Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-09-18 paper/sections/proof-reconstruction.tex
[ paper ] working on structurally smaller arguments.

% -*- root: main.tex -*-
\documentclass[../main.tex]{subfiles}
\begin{document}

@@

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by \citeauthor{sultana2015} in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is the proof
assistant, \Agda. We choose \Agda, but a proof assistant with
support of type theory and similar features like \Agda
(see a description of \Agda in Section
\ref{sec:implementation}) can be used.
Following the diagram in Fig.
\ref{fig:proof-reconstruction-workflow}, the process can begin with
a problem in \CPL that we encode in \TPTP format to be sent as the
input of the \Metis prover.
The algorithms of \Metis in step~3 intend to find
a solution; if they success, \Metis replies a derivation in \TSTP
format of the proof, else the problem is counter feasible.

@@

Finally, we use the proof assistant in step 7 to type-check the
proof and verify its correctness. Next, we only have one answer:
type-checking succeed (i.e., the proof is valid) or type-checking
failed.
If the type-check fails, one of the three actors in the process are
responsible: \Metis, \Athena, or \Agda. We are responsible for
errors committed in the translation from \TSTP to \Agda; lack of the
coverage of the \Metis' inference rules, or bad assumptions about
them. \improvement{Make this statement stronger and clear.}

In the rest of this section, we provide a formal description using
type theory to build definitions and theorems for the functions
necessary to emulate the \Metis' inference rules. We will write the
type judgments with ``$a : A$'' to denote that $a$ has type $A$.
For propositions, we will denote by $φ ≡ ψ$ the syntactic
equivalence.

@@

\subsection{Small Structured Recursion}
\label{ssec:small-structured-recursion}

In the process to emulate \Metis' inference rules in type theory,
we encountered many recursive algorithms with termination problems,
that is,when an algorithm could run forever or we can not guarantee
that it will stop after some point. Such a problem makes impossible
to translate the algorithm in type theory directly.
For that reason, we show a technique that will help us to define
new algorithms based on the previous ones to solve the problem
mentioned.

A recursive function is a \emph{small-structured-recursion}, if the
size of the arguments with which it is defined decrease in
size for each call to the function in the definition~\cite{Bove2005}.
A well-structured function terminates and it can be translated
directly to type theory~\cite{Bove2002}.

Let us consider the following example from an inner function
in the \strip inference.

@@

@@

We reformulate \mysplit function to match with the
new definition after apply the process described above in
Subsection~\ref{ssec:small-structured-recursion}, since
in the recursion calls of \mysplit in Eq.~\ref{eq:split-definition},
the negations, $(¬ φ_{1,2})$ raise a terminating problem.

@@

\emph{Resolve.}

The \resolve rule is a generalization version for resolution
in propositional logic since the formulas are not shaped always as
the theorem expect (see Fig.~\ref{fig:metis-inferences}).
Since our approach is mere syntactic, in order to apply such a
theorem, we perform rearrangements inside the formulas to match with
the pattern expected by the resolution theorem. If this strategy
fails at some point, we could not check the rest of the proof.

Therefore, the best scenario to apply resolution occurs when the
literal $l$ and its negation appear in the exact place in the
formulas as the resolve inference shows.

\begin{figure}
\[%\scalebox{0.9}{
\begin{bprooftree}
\AxiomC{$Γ ⊢ (φ₁ ≡ l ∨ ψ₁)$}
\AxiomC{$Γ ⊢ (φ₂ ≡ ¬ l ∨ ψ₂)$}
\RightLabel{\thmresolve}
\BinaryInfC{$Γ ⊢ \resolve(φ₁, φ₂, l, (ψ₁ ∧ ψ₂))$}
\UnaryInfC{Γ ⊢ ψ₁ ∨ ψ₂}
\end{bprooftree}
%}
\]
% \caption{Resolution in \Metis.}
\label{fig:thm-resolve-best-case}
\end{figure}

To rearrange the propositions in order to match with a specific
pattern for applying a particular theorem like resolution, we have
define the following functions and provide
its respective theorems in \cite{AgdaMetis}.
These rearrangement functions are mainly inspired in
\cite{bohme2010}, for their treatment for conjunction and
disjunction equivalence.

\improvement{provide examples for each of the following functions.}

\begin{itemize}
  \item \name{reorder-∨.} Given two formulas, φ and ψ, both disjunctions, fix the order of the first one by following the syntactically
  order of the second one. If the process succeed (denote by φ ⤳ ψ),
  returns the target formula, ψ, otherwise, keep the first one, φ.
  \begin{align}
  \begin{split}
&\name{reorder-∨}\ :\ \Prop \to \Prop \to \Prop\\
&\name{reorder-∨}(φ, ψ) =
  \begin{cases}
  ψ   & φ ⤳ ψ\\
  ϕ   & \text{otherwise.}
  \end{cases}
  \end{split}
  \end{align}

  \item \name{reorder-∧.} Given two formulas, φ and ψ, both conjunctions, fix the order of the first one by following the order  of the second one.
  \item \name{reorder-∧∨.} Given two formulas, φ and ψ, both in CNF, fix the order of the first one by following the order  of the second one.
\end{itemize}

\begin{equation}
\resolve((φ₁, φ₂, l, ψ)
\end{equation}

\begin{theorem}[\thmresolve] $Γ ⊢ φ₁$, $Γ ⊢ φ₂$,
then $Γ ⊢ \resolve(φ₁, φ₂, l, ψ)$ where $l$ is a literal,
and ψ is the expected result.
\end{theorem}

% \begin{proof}
% \begin{equation*}

% \end{equation*}
% \end{proof}

>>>>

\documentclass[../main.tex]{subfiles}
\begin{document}

@@

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by \citeauthor{sultana2015} in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is the proof
assistant, \Agda. We choose \Agda, but a proof assistant with
support of type theory and similar features like \Agda
(see a description of \Agda in Section
\ref{sec:implementation}) can be used.
Following the diagram in Fig.
\ref{fig:proof-reconstruction-workflow}, the process can begin with
a problem in \CPL that we encode in \TPTP format to be sent as the
input of the \Metis prover.
The algorithms of \Metis in Step~3 intend to find
a solution; if they success, \Metis replies a derivation in \TSTP
format of the proof, else the problem is counter feasible.

@@

Finally, we use the proof checker in Step~7 to type-check the
proof-term. If type-checking succeed, the proof from such a \TSTP
derivation is correct, else if type-checking fails, one of the three
actors in the process could be the responsible.
They are \Metis, \Athena, or \Agda.

We are responsible for errors committed in the translation from
\TSTP to \Agda: bugs in the parsing module, lack of the coverage of
the \Metis' inference rules since we are emulating them, and last
but not least, the pretty print of the proof in \Agda code.

In the rest of this section, we provide a formal description using
type theory to build definitions and theorems for functions
necessary to emulate \Metis' inference rules.
Some conventions about notations. We will write the
type judgments with ``$a : A$'' to denote that $a$ has type
$\abbre{A}$. Types are written using small caps.
For propositions, we will denote by $φ ≡ ψ$, the syntactic
equivalence.

@@

% section section_name (end)
\subsection{Structural Recursion}
\label{ssec:structural-recursion}

In the process to emulate \Metis' inference rules in type theory,
we encountered that some of \Metis' transformation for propositions in
the inference rules were given by \emph{general recursion} functions.
But algorithms based on general recursions can not be translated
directly in type theory since it is not a guarantee they terminate.
Then, since a termination checker in type theory recognizes
\emph{structural recursion functions} as well-founded recursive
functions, we modify our algorithms in order to follow such a
definition mostly taken from \citeauthor{Abel2002}~\cite{Abel2002}.

\begin{definition} A recursive function is structurally recursive if
it calls itself with only \emph{structurally smaller} arguments.
\end{definition}

\begin{definition}
Given an \emph{ordering relation} between $t_1 : T$ and $t_2 : T$,
we say that $t_1$ is structurally smaller than $t_2$ if and only if,
USE \cite{Coquand1992} definition.
\begin{itemize}
  \item $t_1$ and $t_2$ are different constructor for $T$ type.
\end{itemize}
\end{definition}

To convert a general recursion function into a structural recursion
function, for the sake of simplicity, we will follow a technique
found mainly in \cite{Coquand1992,Abel2002,Bove2005}.
It mostly consists of using a \emph{recursor}, an inductive type, to
perform the \emph{induction} over its structure.
A type-checker will recognize structurally smaller arguments by using
the lexicographic order, as an extension of the structural ordering
by the recursor.

Let us consider the following example to exhibit the procedure
mentioned above for an inner function in the \strip inference.

@@

For a more complete treatment of general recursion, and how to
define well-founded and well-structural recursive functions, we refer
the reader to Ref.~\cite{Bove2005}, where Bove and Capettra described
not only general recursion but they also treat mutual recursive
function, among other kinds.

@@

We reformulate \mysplit function to match with the
new definition after apply the process described above in
Subsection~\ref{ssec:structural-recursion}, since
in the recursion calls of \mysplit in Eq.~\ref{eq:split-definition},
the negations, $(¬ φ_{1,2})$ raise a terminating problem.

@@
